SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-04 abs.tex
spell check, editing for grammar, clarity, style, put in some questions preceded by 'djimenez'

Hardware prefetching is an effective technique for hiding cache miss latencies
in modern processor designs.  Prefetcher performance can be characterized by 
two main metrics, which are generally at odds with one another: coverage 
(the improvement in the cache miss metric) and accuracy (the fraction of 
useful prefetches).  An overly aggressive prefetcher may improve coverage, but
may waste so many resources (e.g., bandwidth) that performance is ultimately 
harmed.  An ideal prefetcher would have both very high coverage and accuracy.

@@

In this paper, we introduce Perceptron-based Prefetch Filtering (PPF) as a way
to reject inaccurate candidate prefetches generated by a baseline prefetcher.  
PPF directly increases prefetcher accuracy, and enables more aggressive tuning
of the baseline prefetcher, which in turn leads to increased coverage.  We also
explore a range of features to use to train PPF's perceptron layer.  Our results
show that PPF improves performance on a memory-intensive subset of SPEC CPU 2017
benchmarks by 6.84\% for a single-core configuration, and by 11.9\% for a 4-core
configuration, compared to the baseline prefetcher alone.  We also show that
performance continues to scale as the number of cores sharing a last level cache
increases.

%In this paper, we introduce Perceptron-based Prefetch Filtering to help make
% prefetching decisions accurately. The perceptron layer acts as a check 
%to filter out the unnecessary prefetches recommended by the underlying
%prefetcher.  We have also explored a range of features that can be used to
%train the perceptron layer.  Our results show that perceptron-based filtering
%improves performance on the memory intensive subset of the SPEC 2017 benchmark
%suite by 6.84\% on single-core and by 11.9\% on multi-core traces, as compared
%to a state-of-the art prefetcher.  We also demonstrate that the performance
%gained from using our efficient filter continues to scale as the number of cores sharing a 
%last level cache increases.

>>>>

% djimenez: I have taken out parenthetical terms in favor of explicitly
% mentioning them in the text. some people have a visceral reaction against
% parentheses that delimit things essential to understanding the text.

Hardware prefetching is an effective technique for hiding cache miss latencies
in modern processor designs.  Prefetcher performance can be characterized by
two main metrics that are generally at odds with one another: coverage, {\em
i.e.} the improvement in cache misses; and accuracy, {\em i.e.} the fraction
of useful prefetches.  An overly aggressive prefetcher may improve coverage,
but may waste so many resources, {\em e.g.}, bandwidth, that performance is
harmed.  An ideal prefetcher would have both high coverage and accuracy.

@@

In this paper, we introduce Perceptron-based Prefetch Filtering (PPF) as a way
to reject inaccurate candidate prefetches generated by a baseline prefetcher.
PPF directly increases prefetcher accuracy and enables more aggressive tuning
of the baseline prefetcher, leading to increased coverage.  We also explore a
range of features to use to train PPF's perceptron layer.  PPF improves
performance on a memory-intensive subset of the SPEC CPU 2017 benchmarks by
6.84\% for a single-core configuration, and by 11.9\% for a 4-core
configuration, compared to the baseline prefetcher alone.  We also show that
performance continues to scale as the number of cores sharing a last level
cache increases.

%In this paper, we introduce Perceptron-based Prefetch Filtering to help make
% prefetching decisions accurately. The perceptron layer acts as a check 
%to filter out the unnecessary prefetches recommended by the underlying
%prefetcher.  We have also explored a range of features that can be used to
%train the perceptron layer.  Our results show that perceptron-based filtering
%improves performance on the memory intensive subset of the SPEC CPU 2017 benchmark
%suite by 6.84\% on single-core and by 11.9\% on multi-core traces, as compared
%to a state-of-the art prefetcher.  We also demonstrate that the performance
%gained from using our efficient filter continues to scale as the number of cores sharing a 
%last level cache increases.

### 2018-12-04 back.tex
Completing Methodology, Images
spell check, editing for grammar, clarity, style, put in some questions preceded by 'djimenez'

In this section we discussed the most closely related work to our proposed technique. Other relevant work will be discussed throughout the paper. See section \ref{related}.

	The idea of prefetching has been around ever since Jouppi introduced \textit{instruction stream buffers}. Early prefetchers detected stride access patterns in order to predict future memory requests ~\cite{Smith,Baer,Stride}. These days prefetching mechanisms are more sophisticated, they look into past memory behavior~\cite{Address_Correlated,AMPM}, locality ~\cite{Spatial_Pattern,SMS,Temporal_Instruction_Fetch,Off_Chip,STMS,SMS_JILP}, control-flow speculation ~\cite{BFetch,MTBFetch}, between many other aspects in order to detect complex access patterns in modern applications.

@@

An optimal prefetcher should be able to detect a wide range of memory access patterns. Simple stride prefetching techniques only detect sequences of addresses that differ by a constant value and fail to capture diverse delta patterns. Lookahead prefetchers learn patterns by collecting histories of observed data access patterns, and correlating these with the strides, by considering each new page in a vacumm. Kim textit{et. al.} proposed SPP ~\cite{SPP}, a confidence-based lookahead prefetcher. SPP creates a signature associated to a memory request by compressing the history of accesses. By correlating the signature with future likely delta patterns, SPP learn both simple and complicated memory access patterns quickly. Our proposed technique builds upon SPP, therefore, we now introduce SPP in more detail.

@@

While the basic idea of perceptron based prefetch filtering is applicable to
any look-ahead prefetcher, we developed a practical implementation of the idea
using SPP as our baseline.  Here we describe the basic architecture of SPP.

% djimenez: Paul, is all this stuff accurate? SPP continues to confuse me.

@@

\textbf{Pattern Table:} The Pattern Table is indexed by the signature generated
from the Signature Table.  Pattern Table holds predicted delta patterns and their confidence
estimates.  Each entry indexed by the signature holds up to 4 unique
delta predictions.  This is implemented by making the Pattern Table as a 4-way
associative table.

@@

%<Insert a comprehensive figure about SPP structures>

\begin{figure}
  \begin{center}
  \includegraphics[width=9cm]{figures/SPP_Prefetch_Description.jpg}
    \label{fig:spp_strcture}
  \caption{SPP architecture}
  \end{center}
\end{figure}

%<Insert another detailed picture about SPP data-path flow>
\begin{figure}
  \begin{center}
  \includegraphics[width=9cm]{figures/SPP_Update_Description.jpg}
  \label{fig:spp_update}
  \caption{SPP data-path flow}
  \end{center}
\end{figure}

@@

Moreover, the on-line confidence mechanism introduced in SPP was very
rudimentary.  It was based on taking a ratio C\textsubscript{d} =
C\textsubscript{delta} / C\textsubscript{sig} as explained previously. This
confidence was used to make the decision of whether to prefetch or not to
prefetch; and which level to prefetch.  While this approximation was shown to
work in the original implementation, we believe that a better form of
generalised on-line decision making was possible.  Hence, it was necessary to
build a robust and adaptable learning mechanism to accept / reject the
prefetch suggestions; and to decide the fill level (L2 Cache vs Last Level
Cache).

@@

\subsection{Supervised Learning}

\textit{Perceptron Learning Rule:} In a perceptron-based mechanism, a uniform
perceptron update principle is followed.  The weights need to be updated if
the prediction was wrong or the magnitude of the dot product does not exceed
a certain threshold.  After a certain training period, these weights are
proportional to the probability of a correct prediction.  Thresholding the outcome to a
 magnitude makes sure the perceptrons are trained until a certain level of
confidence and yet they are not over-trained to the given set.  The perceptron 
learning rule will be applied to PPF.

>>>>

In this section we discuss the most closely related work to our proposed
technique. See Section~\ref{related} for other relevant work.

The idea of prefetching begins with Jouppi's \textit{instruction stream
buffers}. Early prefetchers detected stride access patterns in order to
predict future memory requests~\cite{Smith,Baer,Stride}.  Modern prefetching
mechanisms are more sophisticated as they look into past memory
behavior~\cite{Address_Correlated,AMPM},
locality~\cite{Spatial_Pattern,SMS,Temporal_Instruction_Fetch,Off_Chip,STMS,SMS_JILP},
control-flow speculation~\cite{BFetch,MTBFetch}, and many other aspects to
detect complex access patterns in modern applications.

@@

An optimal prefetcher should be able to detect a wide range of memory access
patterns. Simple stride prefetching techniques only detect sequences of
addresses that differ by a constant value and fail to capture diverse delta
patterns. Lookahead prefetchers learn patterns by collecting histories of
observed data access patterns, and correlating these with the strides, by
considering each new page in a vacuum. Kim {\em et. al.} proposed
SPP~\cite{SPP}, a confidence-based lookahead prefetcher. SPP creates a
signature associated to a memory request by compressing the history of
accesses. By correlating the signature with future likely delta patterns, SPP
learn both simple and complicated memory access patterns quickly. Our proposed
technique builds upon SPP. Thus, we describe SPP in more detail.

@@

While the basic idea of perceptron based prefetch filtering is applicable to
any look-ahead prefetcher, we develop a practical implementation of the idea
using SPP as our baseline.  Here we describe the basic architecture of SPP.

@@

\textbf{Pattern Table:} The Pattern Table is indexed by the signature
generated from the Signature Table.  Pattern Table holds predicted delta
patterns and their confidence estimates. Each entry indexed by the signature
holds up to 4 unique delta predictions.  This is implemented by making the
Pattern Table 4-way associative.

@@

\begin{figure}
  \begin{center}
  \includegraphics[width=9cm]{SPP_Prefetch_Description}
    \label{fig:spp_strcture}
  \caption{SPP architecture}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics[width=9cm]{SPP_Update_Description}
  \label{fig:spp_update}
  \caption{SPP data-path flow}
  \end{center}
\end{figure}

@@

Moreover, the on-line confidence mechanism introduced in SPP was very
rudimentary.  It was based on taking a ratio C\textsubscript{d} =
C\textsubscript{delta} / C\textsubscript{sig} as explained previously. This
confidence was used to make the decision of whether to prefetch or not to
prefetch; and which level to prefetch.  While this approximation was shown to
work in the original implementation, we believe that a better form of
generalized on-line decision making was possible.  Hence, it was necessary to
build a robust and adaptable learning mechanism to accept / reject the
prefetch suggestions; and to decide the fill level (L2 Cache vs Last Level
Cache).

@@

\subsection{Perceptron Learning}
\label{sec:Background-Perceptron}
Perceptron learning for microarchitectural prediction was introduced for
branch prediction~\cite{PerceptronPredictor}. Our predictor uses a version of
microarchitectural perceptron prediction known as the ``hashed perceptron''
organization~\cite{HashedPerceptron}. As an abstract idea, a hashed perceptron
predictor hashes several different features into values that index several
distinct tables. Small integer weights are read out from the tables and
summed. If the sum exceeds some threshold, a positive prediction is made, {\em
e.g.} ``predict branch taken'' or ``allow the prefetch.'' Otherwise, a
negative prediction is made. Once the ground truth is known, the weights
corredponding to the prediction are incremented if the outcome was positive,
or decremented if it was negative. This update only occurs if the prediction
was incorrect or if the magnitude of the sum failed to exceed a threshold.
Beyond branch prediction, perceptron learning has been applied to last-level
cache reuse prediction~\cite{Perc_Reuse,Multiperspective}. In this paper, we
apply it for the first time to prefetch filtering.

%\textit{Perceptron Learning Rule:} In a perceptron-based mechanism, a uniform
%perceptron update principle is followed.  The weights need to be updated if
%the prediction was wrong or the magnitude of the dot product does not exceed a
%certain threshold.  After a certain training period, these weights are
%proportional to the probability of a correct prediction.  Thresholding the
%outcome to a
% magnitude makes sure the perceptrons are trained until a certain level of
%confidence and yet they are not over-trained to the given set.  The perceptron
%learning rule will be applied to PPF.

### 2018-12-04 conc.tex
Separating arch and implementation. Filter-Table and Throttler-Filter renaming
spell check, editing for grammar, clarity, style, put in some questions preceded by 'djimenez'

\section{Conclusion}
\label{Conclusion}
In this paper, we introduced neural network based learning for data
prefetching in the form of PPF.  Perceptron acts an orthogonal
throttler to the underlying prefetch engine.  PPF improves
performance by over XX\% over the baseline, which corresponds to YY\%
over the next best performing prefetcher.  
We believe that this is a robust and adaptable technique which can
be used to enhance any existing or upcoming prefetchers.

>>>>

\section{Conclusion}
\label{Conclusion}
In this paper, we introduced perceptron-based learning for data prefetching.
The perceptron filter acts independently as a check on the underlying prefetch
engine.  PPF improves performance by over XX\% over the baseline,
corresponding to YY\% over the next best performing prefetcher.  We believe
that this is a robust and adaptable technique that can be used to enhance any
existing or upcoming prefetchers.

### 2018-12-04 design_arch.tex
spell check, editing for grammar, clarity, style, put in some questions preceded by 'djimenez'

As previously discussed, ultimately it might be profitable to allow a
given prefetcher to speculate as deeply as possible, since often some
useful prefetches are still being generated long after the confidence
of the prefetcher has fallen below the point at which performance
improvements are seen, due to the increasing rate of inaccurate
prefetches.  In order to allow this deep speculation in the
prefetcher, however, the increasing numbers of bad prefetches must be
filtered out.  To this end, we propose to leverage perceptron-based
machine learning as a mechanism to differentiate between potentially
useful deeply speculated prefetches and likely not-useful ones.  This
Perceptron Prefetch Filter (PPF) is placed between the prefetcher and
the prefetch insertion queue, to ensure that predicted bad prefetches
are not inserted into the memory system, 
\pg{As shown in figure whatever...}.

As discussed in Section~\ref{sec:Background-Perceptron}, perceptrons
are useful, light-weight mechanisms to pull together disparate forms
of information and synthesize a decision from them.  In this case, we
will leverage a variety of different relative characteristics of a
given prefetch, such as speculation depth, page address and offset,
into a binary useful/non-useful prediction.  In the remainder of this
section, we discuss the the general structure of PPF for enhancing a
given underlying prefetcher.  We will show that PPF is a generalized
prefetch filtering mechanism which may adapted to any prefetcher with
appropriate modifications.
% Gino: Are these modifications minimal?
% EB: That's a good question :P. Edited to 'appropriate'.

@@

%Gino: repharse below?
The Perceptron Filter is organized as tables of perceptron weights,
with a separate table of weights for each feature. For a design using 
`N' number of features, it's weights will be arranged in N different 
tables. Each feature's value is used to index into the corresponding table.
Since each table gets indexed independently, different number of bits can 
be used for indexing, thus giving each table a different depth.
The variable indexing can be determined by studying the features and
fine-tuned empirically to achieve a good accuracy vs hardware overhead
trade-off.  Exact details of our PPF implementation are mentioned 
under ``Area Overheads'' in Section~\ref{Method-Overheads}.

@@

A single entry in the table corresponds to a perceptron weight.  Each
weight is a 5 bit counter - saturating at -16 and +15 and initialized
to zero at the beginning of a program's execution. It was empirically 
observed that 5-bits of indexing achieves a good trade-off between 
hardware overheads and accuracy. 
%\pg{the fact that its nine is again implementation details}
% [EB] Resolved

% \pg{The word ChampSim should not appear in this part of the paper at
%   all.  Remember you are describing how to build this in the general
%   case...} 
% [EB] Resolved
\textbf{Inference}\newline The prefetcher gets triggered on every 
L2 Cache demand access.  At
that point, the base prefetcher has the option to perform a prefetch
-- if it does, then it has a choice of how many cache lines to
prefetch.  The suggested prefetches can be either placed in the L2
cache or the last-level cache based on the prefetcher's internal
confidence mechanism.

When the base prefetcher is triggered, it begins to suggest candidates
for prefetching.  All suggested prefetches are passed through the
perceptron filter to decide if they are qualified for final
prefetching.  The perceptron decides whether to prefetch a candidate
by looking at the microarchitectural state \textit{i.e.} the features,
at that instant.  Each feature is hashed to form an index into a table
of up to 4096 entries dedicated for that feature.

Once all the weights are retrieved, they are summed and compared to a
preset threshold, PERC\_ THRESHOLD\_LO.  Only the prefetch candidates
with perceptron sum higher than the threshold qualify for prefetching.
The prefetches that qualify through the perceptron stage are recorded
in the ``Prefetch Table''. The prefetch table is a 1024-entry, direct
mapped structure that contains all metadata required to re-index the 
perceptron entries for training.  
When the feedback of the current prefetched line is available at a
later stage, the stored data is used to train the perceptron.
%\pg{this discussion below has redundancy and doesn't flow well.
%  Rework it}

Each feature's confidence contributes to the final decision on 
whether to prefetch a cache line, with the summation of all features' 
confidences dentoting the overall confidence for a prefetch. 
The final sum is compared against two thresholds: 
 PERC\_THRESHOLD\_HI and  PERC\_ THRESHOLD\_ LO to decide the
 placement of the data in the cache heirarchy. Prefetches that 
have a confidence above the higher threshold are placed into 
the L2 cache since the observed features strongly recommend  
the prefetch. A prefetch that results in a confidence between 
PERC\_ THRESHOLD\_ LO and PERC\_ THRESHOLD\_ HI are placed in 
the larger last level cache, as the filter is moderately confident of 
the cache line's use, but seeks to reduce pollution in the smaller 
L2 cache. Finally, the remaining prefetches that are below the 
lower threshold are not prefetched due to a lack of correlation 
between the current observed features and the predicction from 
the base prefetch engine.

@@

In addition to the prefetch table mentioned above, PPF also maintains
a ``Reject Table.''  The reject table is a 1024-entry deep
direct-mapped table.  If a prefetch suggestion is rejected by the
perceptron layer, it is logged into the reject table.  The table is
used to train the perceptron to 
avoid false negatives \textit{i.e.}, cases where prediction was to
reject the prefetch but the prefetch would have been useful.

@@

\textbf{Training}\newline In the prefetching environment, feedback for
a prefetch is received whenever there is an eviction or a demand
access from the L2 Cache.  This action triggers training of the base
prefetcher as well as the perceptron layer.  Training the perceptrons
involves accessing the metadata that was stored in Prefetch Table or
the Reject Table.  The cache lines address of the block being trained
is used to access a table, using 10-bits of the address for indexing
and another 6-bits to perform tag matching.  Once the values of the
features at the time of prefetching is available, it is
used to index into the perceptron weights table.

If the demanded block that triggers the training was tagged as a valid
prefetch in the prefetch table, then the earlier prefetch prediction
was correct.  In that case the perceptron weights are incremented by 1
if the predicted sum does not cross a pre-defined threshold. These
training thresholds are introduced to avoid overfitting of the
perceptron weights to the given program behaviour. These thresholds
are referred to as POS\_UPDATE\_THRESHOLD ($\theta_p$) and
NEG\_UPDATE\_THRESHOLD ($\theta_n$), respectively for the positive and
negative values of training saturation.

@@

\subsection{Optimizing PPF for a Given Prefetcher}
\label{Arch-Generalizing}
The above discussion of PPF shows that it is highly modular and can be
adapted to be used over any base prefecher for increased prefetch
accuracy.  In general, only two hooks are required between PPF and the
baseline prefetcher. The first is to make sure that all the prefetch
candidates of the prefetcher pass through the perceptron filter and if
qualified, the metadata for perceptron indexing be stored. The second
is needed when the feedback of a prior prefetch is available in form
of a subsequent demand hit or cache eviction. In that case, the stored
metadata needs to be retrieved to update the state of the perceptrons.

@@

\item \textbf{Feature Selection:}  Perceptrons essentially normalize
  contributions from different features to get a single sum representing the
  final confidence. So the perceptron learning can be as good as the set of 
  features chosen. Interestingly, this is what makes perceptron learning 
  scalable, as it can easily learn to incorporate newer information in 
  form of new features. Some of the features we developed use information 
  derived directly from program execution, agnostic to the baseline 
  prefetcher. Beyond that, the feature set can be expanded to convey 
  any useful information from the prefetcher to the perceptron filter.

>>>>

As previously discussed, ultimately it might be profitable to allow a given
prefetcher to speculate as deeply as possible, since often some useful
prefetches are generated long after the confidence of the prefetcher has
fallen below the point at which performance improvements are seen due to the
increasing rate of inaccurate prefetches.  However, to allow this deep
speculation in the prefetcher the increasing numbers of bad prefetches must be
filtered out.  Thus, we propose to leverage perceptron-based learning as a
mechanism to differentiate between potentially useful deeply speculated
prefetches and likely not-useful ones.  This Perceptron Prefetch Filter (PPF)
is placed between the prefetcher and the prefetch insertion queue, to ensure
that predicted bad prefetches are not inserted into the memory system.
\pg{As shown in figure whatever...}.

As discussed in Section~\ref{sec:Background-Perceptron}, perceptron learning
is a light-weight mechanism to pull together disparate forms of information
and synthesize a decision from them.  In this case, we leverage a variety of
different relative characteristics of a given prefetch, such as speculation
depth, page address and offset, into a binary useful/non-useful prediction.
In the remainder of this section, we discuss the the general structure of PPF
for enhancing a given underlying prefetcher.  We will show that PPF is a
generalized prefetch filtering mechanism that may be adapted to any prefetcher
with appropriate modifications.

% Gino: Are these modifications minimal?
% EB: That's a good question :P. Edited to 'appropriate'.

@@

%Gino: rephrase below?
The Perceptron Filter is organized as tables of perceptron weights, with a
separate table of weights for each feature. For a design using `N' number of
features, its weights will be arranged in $N$ different tables. Each feature's
value is used to index into the corresponding table.  Since each table is
indexed independently, different numbers of bits can be used for indexing,
thus giving each table a different size.  The variable indexing can be
determined by studying the features and fine-tuning them empirically to
achieve a good accuracy versus hardware overhead trade-off.  Exact details of
our PPF implementation are given under ``Area Overheads'' in
Section~\ref{Method-Overheads}.

@@

A single entry in the table corresponds to a perceptron weight.  Each weight
is a 5 bit counter saturating at -16 and +15 and initialized to zero at the
beginning of a program's execution. We observed empirically that 5 bits of
index achieves a good trade-off between hardware overheads and accuracy.

%\pg{the fact that its nine is again implementation details}
% [EB] Resolved

% \pg{The word ChampSim should not appear in this part of the paper at
%   all.  Remember you are describing how to build this in the general
%   case...} 
% [EB] Resolved
\textbf{Inference}\newline The prefetcher is triggered on every L2 cache
demand access.  At that point, the base prefetcher has the option to perform a
prefetch -- if it does, then it has a choice of how many cache lines to
prefetch.  The suggested prefetches can be either placed in the L2 cache or
the last-level cache based on the prefetcher's internal confidence mechanism.

When the base prefetcher is triggered, it begins to suggest candidates for
prefetching.  All suggested prefetches are passed through the perceptron
filter to decide whether they are qualified for final prefetching.  The filter
decides whether to prefetch a candidate by looking at the microarchitectural
state, \textit{i.e.} the features, at that instant.  Each feature is hashed to
form an index into a table of up to 4,096 entries dedicated to that feature.

Once all the weights are retrieved, they are summed and compared to a preset
threshold, PERC\_ THRESHOLD\_LO.  Only prefetch candidates with perceptron sum
higher than the threshold qualify for prefetching.  The prefetches that
qualify through the perceptron stage are recorded in the ``Prefetch Table''.
The prefetch table is a 1,024-entry, direct mapped structure that contains all
metadata required to re-index the perceptron entries for training.  When the
feedback of the current prefetched line is available at a later stage, the
stored data is used to train the perceptron.

%\pg{this discussion below has redundancy and doesn't flow well.
%  Rework it}

Each feature's confidence contributes to the final decision of whether to
prefetch a cache line, with the summation of all features' confidences
denoting the overall confidence for a prefetch.  The final sum is compared
against two thresholds: PERC\_THRESHOLD\_HI and  PERC\_ THRESHOLD\_ LO to
decide the placement of the data in the cache hierarchy. Prefetches that have
a confidence above the higher threshold are placed into the L2 cache since the
observed features strongly recommend  the prefetch. A prefetch that results in
a confidence between PERC\_ THRESHOLD\_ LO and PERC\_ THRESHOLD\_ HI are
placed in the larger last level cache, as the filter is moderately confident
of the cache line's use, but seeks to reduce pollution in the smaller L2
cache. Finally, the remaining prefetches that are below the lower threshold
are not prefetched due to a lack of correlation between the current observed
features and the prediction from the base prefetch engine.

@@

In addition to the prefetch table mentioned above, PPF also maintains a
``Reject Table.''  The reject table is a 1,024-entry direct-mapped table.
If a prefetch suggestion is rejected by the perceptron layer, it is logged
into the reject table.  The table is used to train the perceptron to avoid
false negatives \textit{i.e.}, cases where prediction was to reject the
prefetch but the prefetch would have been useful.

@@

\textbf{Training}\newline In the prefetching environment, feedback for a
prefetch is received whenever there is an eviction or a demand access from the
L2 cache.  This action triggers training of the base prefetcher as well as the
perceptron layer.  Training the perceptrons involves accessing the metadata
that was stored in Prefetch Table or the Reject Table.  The cache lines
address of the block being trained is used to access a table, using 10 bits of
the address for indexing and another 6 bits to perform tag matching.  Once the
values of the features at the time of prefetching is available, it is used to
index into the perceptron weights table.

If the demanded block that triggers the training was tagged as a valid
prefetch in the prefetch table, then the earlier prefetch prediction was
correct.  In that case the perceptron weights are incremented by 1 if the
predicted sum does not cross a pre-defined threshold. These training
thresholds are introduced to avoid overfitting of the perceptron weights to
the given program behavior. These thresholds are referred to as
POS\_UPDATE\_THRESHOLD ($\theta_p$) and NEG\_UPDATE\_THRESHOLD ($\theta_n$),
respectively for the positive and negative values of training saturation.

@@

\subsection{Optimizing PPF for a Given Prefetcher}
\label{Arch-Generalizing}
The above discussion of PPF shows that it is highly modular and can be adapted
to be used over any base prefetcher for increased prefetch accuracy.  In
general, only two hooks are required between PPF and the baseline prefetcher.
The first is to make sure that all the prefetch candidates of the prefetcher
pass through the perceptron filter and if qualified, the metadata for
perceptron indexing be stored. The second is needed when the feedback of a
prior prefetch is available in form of a subsequent demand hit or cache
eviction. In that case, the stored metadata needs to be retrieved to update
the state of the perceptrons.

@@

\item \textbf{Feature Selection:}  Perceptrons essentially normalize
  contributions from different features to get a single sum representing the
  final confidence. So the perceptron learning can be as good as the set of 
  features chosen. Interestingly, this is what makes perceptron learning 
  scalable, as it can easily learn to incorporate newer information in 
  the form of new features. Some of the features we developed use information 
  derived directly from program execution, agnostic to the baseline 
  prefetcher. Beyond that, the feature set can be expanded to convey 
  any useful information from the prefetcher to the perceptron filter.

### 2018-12-04 design_impl.tex
spell check, editing for grammar, clarity, style, put in some questions preceded by 'djimenez'

\subsection{Features used by Perceptron}
\label{Impl-Features}
Here we discuss the various features that correlate the prefetching
decision with the program behavior. All the features we used can be 
derived from the information available in the L2 Cache access stream.
Our feature selction involved searching over a large space of 
relevant perceptron features. Using the statistical methodology 
outlined in Section~\ref{Method-Features}, we pruned the feature 
set to a minimal yet relevant set of features
% \pg{again you want to get at why
% these were chosen.  Also mention that you searched over a large
% space of possible features and these were the best.}

% [EB] Tried addressing. Does it look reduntant when read after 'Feature Selection'
% bullet from the previous section?

\begin{itemize}
\item \textbf{Base Address}: The base address of the demand 
access that triggers the prefetch coresponds to a stream of 
future accesses that SPP and the PPF have seen before. 
Therefore, PPF will correlate the past behavior of this address 
to the confidence of a prefetch.

@@

\item \textbf{Program Counter XOR Delta}: This feature tells us if a
  given PC favours particular value(s) of delta.  As noted earlier,
  while the PC alone does not convey useful information, this
  hash resolves the PC into different values based on the tendency of
  that PC to favour a certain delta.  Thus, the dynamic nature of
  different instances of the same memory access instruction is 
  captured here.
 % GAC: Seems more predictable than dynamic if its capturing 
 % what is essentially a streaming access pattern. Not quite sure how to rephrase,
 % if I'm correct

 % GAC: Again having trouble phrasing the below point. I was attempting to say
 % that despite SPP having low confidence due to speculation, PPF may 
 % reinforce prefetches with lower confidence if SPP has had a tendancy to be 
 % correct in the past.
\item \textbf{Confidence}: The integer confidence, on a scale of 0 to
  100, that was used in the original SPP design. While the original 
  confidence does not directly make the decision to prefetch,
  PPF correlates it to the correctness of a proposed prefetch.
  While the original SPP may have dismissed a prefetch due to running
  further into speculation, PPF can use the original confidence as indicator
  of not only when prefetches become less confident, but also how likely
  a low confident speculation is correct in the context of other features.

@@

%%%%%%% Below are the features rejected when downsizing 14 -> 9
% \item \textbf{Current Signature}: The 12-bit signature used by SPP
%   to index into the Pattern Table
% \item \textbf{Delta}: The signed version of the difference in the
%   block address that triggers the prefetch and the block address of
%   the data being prefetched
% \item \textbf{PC}: Program Counter of the Load instruction that
%   triggers that particular prefetch
% \item \textbf{Base address XOR Delta}: If any particular value(s) of
%   delta are highly favoured by a given address which triggered the
%   prefetch, this composite feature can capture that information
% \item \textbf{Cache line XOR Delta}: qThis feature captures if any
%   particular delta is highly favoured by a given cache line address
%\item \textbf{Confidence XOR Delta}: This gives a combined insight
%  into that fact that certain delta-confidence pairqs predicted by
%  SPP might correlate more with the prefetch outcome
%%%%%%%%%%%%%%%%%%%%%%

@@

Some of the composite features are derived from simple hashing (XOR)
of two primary features.  There is always a question of usefulness of
such composite features and the new information added.  We justify the
choice of each feature by quantifying the contribution made towards
predicting prefetch behaviour, in Section \ref{Method-Features}.
Finally, as noted above, each feature indexes into its independent
entry of perceptron weights.

>>>>

\subsection{Features used by Perceptron}
\label{Impl-Features}
Here we discuss the various features that correlate the prefetching
decision with the program behavior. All the features we used can be 
derived from the information available in the L2 Cache access stream.
Our feature selection involved searching over a large space of 
relevant perceptron features. Using the statistical methodology 
outlined in Section~\ref{Method-Features}, we pruned the feature 
set to a minimal yet relevant set of features
% \pg{again you want to get at why
% these were chosen.  Also mention that you searched over a large
% space of possible features and these were the best.}

% [EB] Tried addressing. Does it look redundant when read after 'Feature Selection'
% bullet from the previous section?

\begin{itemize}
\item \textbf{Base Address}: The base address of the demand 
access that triggers the prefetch corresponds to a stream of 
future accesses that SPP and the PPF have seen before. 
Therefore, PPF will correlate the past behavior of this address 
to the confidence of a prefetch.

@@

\item \textbf{Program Counter XOR Delta}: This feature tells us if a
  given PC favors particular value(s) of delta.  As noted earlier,
  while the PC alone does not convey useful information, this
  hash resolves the PC into different values based on the tendency of
  that PC to favor a certain delta.  Thus, the dynamic nature of
  different instances of the same memory access instruction is 
  captured here.
 % GAC: Seems more predictable than dynamic if its capturing 
 % what is essentially a streaming access pattern. Not quite sure how to rephrase,
 % if I'm correct

 % GAC: Again having trouble phrasing the below point. I was attempting to say
 % that despite SPP having low confidence due to speculation, PPF may 
 % reinforce prefetches with lower confidence if SPP has had a tendency to be 
 % correct in the past.
\item \textbf{Confidence}: The integer confidence, on a scale of 0 to
  100, that was used in the original SPP design. While the original 
  confidence does not directly make the decision to prefetch,
  PPF correlates it to the correctness of a proposed prefetch.
  While the original SPP may have dismissed a prefetch due to running
  further into speculation, PPF can use the original confidence as indicator
  of not only when prefetches become less confident, but also how likely
  a low confident speculation is correct in the context of other features.

@@

%%%%%%% Below are the features rejected when downsizing 14 -> 9
% \item \textbf{Current Signature}: The 12-bit signature used by SPP
%   to index into the Pattern Table
% \item \textbf{Delta}: The signed version of the difference in the
%   block address that triggers the prefetch and the block address of
%   the data being prefetched
% \item \textbf{PC}: Program Counter of the Load instruction that
%   triggers that particular prefetch
% \item \textbf{Base address XOR Delta}: If any particular value(s) of
%   delta are highly favored by a given address which triggered the
%   prefetch, this composite feature can capture that information
% \item \textbf{Cache line XOR Delta}: this feature captures if any
%   particular delta is highly favored by a given cache line address
%\item \textbf{Confidence XOR Delta}: This gives a combined insight
%  into that fact that certain delta-confidence pairs predicted by
%  SPP might correlate more with the prefetch outcome
%%%%%%%%%%%%%%%%%%%%%%

@@

Some of the composite features are derived from simple hashing (XOR)
of two primary features.  There is always a question of usefulness of
such composite features and the new information added.  We justify the
choice of each feature by quantifying the contribution made towards
predicting prefetch behavior, in Section \ref{Method-Features}.
Finally, as noted above, each feature indexes into its independent
entry of perceptron weights.
